---
title: "How significant are kills in the NBA?"
author: "Shashwat Mishra"
date: "2026-02-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(hoopR)
library(tidyr)
```

# Kills in the NBA: Testing Sean Miller's theory of kills & win percentage

There really are a lot of new things you learn in college outside of class. Sometimes, it can happen at a basketball game. After the departure of Texas phenom Tre Johnson to the NBA, and subsequently, coach Rodney Terry (after a miraculous entry into the tournament and proceeding 1st round exit), I was excited to see Sean Miller take over.

After going to the Longhorns' season opener, the pregame video was about a stat I'd never heard before: kills. A kill is simple, 3 defensive stops in a row, be it by missed shot + defensive rebound, offensive foul, steal, etc. Miller proclaimed that reaching the area around 7+ kills brings win probability of the game up to 90%.

This project just look at how kills translate to the NBA, a league dominated by star-studded offenses.

## Part 1: Testing Scope: Using a smaller sample (Dallas Mavericks 2024-2025)

### 1. Load NBA play-by-play data for the 2024-25 season

```{r}
season24_25 <- load_nba_pbp(
  seasons = 2025
)
```

### 2. Look at only Mavs games

```{r}
mavs_2025 <- season24_25 %>%
  filter(home_team_abbrev == "DAL" | away_team_abbrev == "DAL")
```

### 3. Analyzing Defensive Stops by text data

```{r}
mavs_2025 <- mavs_2025 %>%
  arrange(game_id, period, desc(clock_seconds)) %>%
  mutate(
    mavs_on_defense = case_when(
      home_team_abbrev == "DAL" & team_id == away_team_id ~ TRUE,
      away_team_abbrev == "DAL" & team_id == home_team_id ~ TRUE,
      TRUE ~ FALSE
    ),

    opp_turnover = mavs_on_defense &
      grepl(
      "Turnover|Offensive Foul|Violation",
      type_text,
      ignore.case = TRUE
    ),

    opp_missed_shot = mavs_on_defense &
      grepl("misses", text),

    mavs_def_reb = !mavs_on_defense &
      type_text == "Defensive Rebound"
  )

#Organize by game in order
mavs_2025 <- mavs_2025 %>%
  arrange(
    game_id,
    period_number,
    desc(clock_minutes),
    desc(clock_seconds)
  )

mavs_2025 <- mavs_2025 %>%
  mutate(
    opp_scored = mavs_on_defense & scoring_play
  )

```

### 4. Defensive Stops

```{r}
mavs_2025 <- mavs_2025 %>%
  mutate(
    defensive_stop = case_when(
      opp_turnover ~ 1,
      opp_missed_shot & lead(mavs_def_reb, default = FALSE) ~ 1,
      TRUE ~ 0
    )
  )
```

### 5. Making a possession index

```{r}
# Create possession index
mavs_2025 <- mavs_2025 %>%
  arrange(
    game_id,
    period_number,
    desc(clock_minutes),
    desc(clock_seconds)
  ) %>%
  mutate(
    possession_change = case_when(
      opp_turnover ~ 1,
      mavs_def_reb ~ 1,
      scoring_play ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(game_id) %>%
  mutate(
    possession = cumsum(possession_change)
  ) %>%
  ungroup()

```

### 6. Kills logic + merging onto original data

```{r}
stops_by_possession <- mavs_2025 %>%
  group_by(game_id, possession) %>%
  summarise(
    stop = as.integer(any(defensive_stop == 1)),
    .groups = "drop"
  )

kills_by_possession <- stops_by_possession %>%
  arrange(game_id, possession) %>%
  group_by(game_id) %>%
  mutate(
    stop_streak = ifelse(
      stop == 1,
      ave(stop, cumsum(stop == 0), FUN = cumsum),
      0
    ),
    kill = ifelse(stop_streak %% 3 == 0 & stop_streak > 0, 1, 0)
  ) %>%
  ungroup()


kills_per_game <- kills_by_possession %>%
  group_by(game_id) %>%
  summarise(
    kills = sum(kill),
    .groups = "drop"
  )

```

```{r}
# Sanity Check
kills_by_possession %>%
  filter(game_id == 401703376) %>%
  select(possession, stop, stop_streak, kill)
```

### 7. Merging with Mavs wins

```{r}
mavs_games <- load_nba_team_box(seasons = 2025) %>%
  filter(team_abbreviation == "DAL")

# Merge kills with results
mavs_kills_wins <- kills_per_game %>%
  left_join(mavs_games, by = "game_id")

```

```{r}
kills_per_game <- kills_by_possession %>%
  group_by(game_id) %>%
  summarise(
    kills = sum(kill),
    .groups = "drop"
  )

nrow(kills_per_game)

```

```{r}
mavs_kills_games <- mavs_games %>%
  left_join(
    kills_per_game,
    by = "game_id"
  ) %>%
  mutate(
    kills = replace_na(kills, 0)
  )
#count total number of kills
mavs_kills_games %>%
  summarise(total_kills = sum(kills))

mavs_kills_games <- mavs_kills_games %>%
  mutate(win_num = as.integer(team_winner))
```

### 8. Logistic Regression Model

```{r}
kills_model <- glm(
  team_winner ~ kills,
  data = mavs_kills_games,
  family = binomial
)

summary(kills_model)
```

### 9. Graphing the relationship between kills and win percentage

```{r}
mavs_kill_summary <- mavs_kills_games %>%
  group_by(kills) %>%
  summarise(
    
    
    win_rate = mean(win_num),
    n = n(),
    .groups = "drop"
  )

ggplot(mavs_kill_summary, aes(x = kills, y = win_rate)) +
  geom_point(size = 3) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Mavericks Win Rate by Number of Kills (2024–25)",
    x = "Number of Kills",
    y = "Win Percentage"
  ) +
  theme_minimal()

```


- home & away team kills, does one team has a lot of kills, does the other team have as many or significantly less.

- find correlation with other defensive metrics (BPG, SPG, etc)

- model with DRTG instead of kills to see difference (compare and see which is better)

- look at the distribution of kills per game, see if there are outliers, and how many games have 7+ kills (the threshold Miller mentioned) and how many of those were wins.

## Moving into the NBA: Expanding the scope to all teams and seasons

```{r}
pbp_2025 <- season24_25 %>%
  arrange(game_id, period_number, desc(clock_minutes), desc(clock_seconds)) %>%
  mutate(
    defending_team_id = case_when(
      team_id == home_team_id ~ away_team_id,
      team_id == away_team_id ~ home_team_id,
      TRUE ~ NA_integer_
    )
  )


```



