---
title: "How significant are kills in the NBA?"
author: "Shashwat Mishra"
date: "2026-02-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(hoopR)
library(tidyr)
```

# Kills in the NBA: Testing Sean Miller's theory of kills & win percentage

There really are a lot of new things you learn in college outside of class. Sometimes, it can happen at a basketball game. After the departure of Texas phenom Tre Johnson to the NBA, and subsequently, coach Rodney Terry (after a miraculous entry into the tournament and proceeding 1st round exit), I was excited to see Sean Miller take over.

After going to the Longhorns' season opener, the pregame video was about a stat I'd never heard before: kills. A kill is simple, 3 defensive stops in a row, be it by missed shot + defensive rebound, offensive foul, steal, etc. Miller proclaimed that reaching the area around 7+ kills brings win probability of the game up to 90%.

This project just look at how kills translate to the NBA, a league dominated by star-studded offenses.

## Part 1: Testing Scope: Using a smaller sample (Dallas Mavericks 2024-2025)

### 1. Load NBA play-by-play data for the 2024-25 season

```{r}
season24_25 <- load_nba_pbp(
  seasons = 2025
)
```

### 2. Look at only Mavs games

```{r}
mavs_2025 <- season24_25 %>%
  filter(home_team_abbrev == "DAL" | away_team_abbrev == "DAL")
```

### 3. Analyzing Defensive Stops by text data

```{r}
mavs_2025 <- mavs_2025 %>%
  arrange(game_id, period, desc(clock_seconds)) %>%
  mutate(
    mavs_on_defense = case_when(
      home_team_abbrev == "DAL" & team_id == away_team_id ~ TRUE,
      away_team_abbrev == "DAL" & team_id == home_team_id ~ TRUE,
      TRUE ~ FALSE
    ),

    opp_turnover = mavs_on_defense &
      grepl(
      "Turnover|Offensive Foul|Violation",
      type_text,
      ignore.case = TRUE
    ),

    opp_missed_shot = mavs_on_defense &
      grepl("misses", text),

    mavs_def_reb = !mavs_on_defense &
      type_text == "Defensive Rebound"
  )

#Organize by game in order
mavs_2025 <- mavs_2025 %>%
  arrange(
    game_id,
    period_number,
    desc(clock_minutes),
    desc(clock_seconds)
  )

mavs_2025 <- mavs_2025 %>%
  mutate(
    opp_scored = mavs_on_defense & scoring_play
  )

```

### 4. Defensive Stops

```{r}
mavs_2025 <- mavs_2025 %>%
  mutate(
    defensive_stop = case_when(
      opp_turnover ~ 1,
      opp_missed_shot & lead(mavs_def_reb, default = FALSE) ~ 1,
      TRUE ~ 0
    )
  )
```

### 5. Making a possession index

```{r}
# Create possession index
mavs_2025 <- mavs_2025 %>%
  arrange(
    game_id,
    period_number,
    desc(clock_minutes),
    desc(clock_seconds)
  ) %>%
  mutate(
    possession_change = case_when(
      opp_turnover ~ 1,
      mavs_def_reb ~ 1,
      scoring_play ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(game_id) %>%
  mutate(
    possession = cumsum(possession_change)
  ) %>%
  ungroup()

```

### 6. Kills logic + merging onto original data

```{r}
stops_by_possession <- mavs_2025 %>%
  group_by(game_id, possession) %>%
  summarise(
    stop = as.integer(any(defensive_stop == 1)),
    .groups = "drop"
  )

kills_by_possession <- stops_by_possession %>%
  arrange(game_id, possession) %>%
  group_by(game_id) %>%
  mutate(
    stop_streak = ifelse(
      stop == 1,
      ave(stop, cumsum(stop == 0), FUN = cumsum),
      0
    ),
    kill = ifelse(stop_streak %% 3 == 0 & stop_streak > 0, 1, 0)
  ) %>%
  ungroup()


kills_per_game <- kills_by_possession %>%
  group_by(game_id) %>%
  summarise(
    kills = sum(kill),
    .groups = "drop"
  )

```

```{r}
# Sanity Check
kills_by_possession %>%
  filter(game_id == 401703376) %>%
  select(possession, stop, stop_streak, kill)
```

### 7. Merging with Mavs wins

```{r}
mavs_games <- load_nba_team_box(seasons = 2025) %>%
  filter(team_abbreviation == "DAL")

# Merge kills with results
mavs_kills_wins <- kills_per_game %>%
  left_join(mavs_games, by = "game_id")

```

```{r}
kills_per_game <- kills_by_possession %>%
  group_by(game_id) %>%
  summarise(
    kills = sum(kill),
    .groups = "drop"
  )

nrow(kills_per_game)

```

```{r}
mavs_kills_games <- mavs_games %>%
  left_join(
    kills_per_game,
    by = "game_id"
  ) %>%
  mutate(
    kills = replace_na(kills, 0)
  )
#count total number of kills
mavs_kills_games %>%
  summarise(total_kills = sum(kills))

mavs_kills_games <- mavs_kills_games %>%
  mutate(win_num = as.integer(team_winner))
```

### 8. Logistic Regression Model

```{r}
kills_model <- glm(
  team_winner ~ kills,
  data = mavs_kills_games,
  family = binomial
)

summary(kills_model)
```

### 9. Graphing the relationship between kills and win percentage

```{r}
mavs_kill_summary <- mavs_kills_games %>%
  group_by(kills) %>%
  summarise(
    
    
    win_rate = mean(win_num),
    n = n(),
    .groups = "drop"
  )

ggplot(mavs_kill_summary, aes(x = kills, y = win_rate)) +
  geom_point(size = 3) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Mavericks Win Rate by Number of Kills (2024–25)",
    x = "Number of Kills",
    y = "Win Percentage"
  ) +
  theme_minimal()

```


- home & away team kills, does one team has a lot of kills, does the other team have as many or significantly less.

- find correlation with other defensive metrics (BPG, SPG, etc)

- model with DRTG instead of kills to see difference (compare and see which is better)

- look at the distribution of kills per game, see if there are outliers, and how many games have 7+ kills (the threshold Miller mentioned) and how many of those were wins.

## Moving into the NBA: Expanding the scope to all teams and seasons

```{r}
# ============================================================
# Kills in the NBA: Full League Analysis (2024-25 + Multi-Season)
# ============================================================

library(dplyr)
library(ggplot2)
library(ggthemes)
library(hoopR)
library(tidyr)
library(lme4)       # mixed effects models
library(scales)
library(purrr)

# ============================================================
# HELPER FUNCTION: Compute kills for one team in one game
# ============================================================

compute_kills_for_team <- function(pbp_data, team_abbrev) {
  team_games <- pbp_data %>%
    filter(home_team_abbrev == team_abbrev | away_team_abbrev == team_abbrev)
  
  team_games <- team_games %>%
    arrange(game_id, period_number, desc(clock_minutes), desc(clock_seconds)) %>%
    mutate(
      team_on_defense = case_when(
        home_team_abbrev == team_abbrev & team_id == away_team_id ~ TRUE,
        away_team_abbrev == team_abbrev & team_id == home_team_id ~ TRUE,
        TRUE ~ FALSE
      ),
      opp_turnover = team_on_defense &
        grepl("Turnover|Offensive Foul|Violation", type_text, ignore.case = TRUE),
      opp_missed_shot = team_on_defense & grepl("misses", text),
      team_def_reb = !team_on_defense & type_text == "Defensive Rebound"
    ) %>%
    mutate(
      defensive_stop = case_when(
        opp_turnover ~ 1,
        opp_missed_shot & lead(team_def_reb, default = FALSE) ~ 1,
        TRUE ~ 0
      ),
      possession_change = case_when(
        opp_turnover ~ 1,
        team_def_reb ~ 1,
        scoring_play ~ 1,
        TRUE ~ 0
      )
    ) %>%
    group_by(game_id) %>%
    mutate(possession = cumsum(possession_change)) %>%
    ungroup()
  
  stops_by_possession <- team_games %>%
    group_by(game_id, possession) %>%
    summarise(stop = as.integer(any(defensive_stop == 1)), .groups = "drop")
  
  kills_by_possession <- stops_by_possession %>%
    arrange(game_id, possession) %>%
    group_by(game_id) %>%
    mutate(
      stop_streak = ifelse(
        stop == 1,
        ave(stop, cumsum(stop == 0), FUN = cumsum),
        0
      ),
      kill = ifelse(stop_streak %% 3 == 0 & stop_streak > 0, 1, 0)
    ) %>%
    ungroup()
  
  kills_by_possession %>%
    group_by(game_id) %>%
    summarise(kills = sum(kill), team_abbrev = team_abbrev, .groups = "drop")
}

# ============================================================
# PART 1: 2024-25 Full League Analysis
# ============================================================

season_pbp <- load_nba_pbp(seasons = 2025)

all_teams <- unique(c(season_pbp$home_team_abbrev, season_pbp$away_team_abbrev))
all_teams <- all_teams[!is.na(all_teams)]

# Compute kills for all teams (takes a few minutes)
all_kills <- map_dfr(all_teams, function(tm) {
  message("Processing: ", tm)
  compute_kills_for_team(season_pbp, tm)
})

# Load team box scores for all teams
team_box <- load_nba_team_box(seasons = 2025)

# Join kills with results
kills_wins <- team_box %>%
  left_join(all_kills, by = c("game_id", "team_abbreviation" = "team_abbrev")) %>%
  mutate(
    kills = replace_na(kills, 0),
    win_num = as.integer(team_winner)
  )

# ---- Opponent kills per game (kills allowed) ----
# Merge home and away kills to get both sides
home_kills <- all_kills %>%
  rename(home_team_abbrev = team_abbrev, home_kills = kills)
away_kills <- all_kills %>%
  rename(away_team_abbrev = team_abbrev, away_kills = kills)

game_kills <- season_pbp %>%
  distinct(game_id, home_team_abbrev, away_team_abbrev) %>%
  left_join(home_kills, by = c("game_id", "home_team_abbrev")) %>%
  left_join(away_kills, by = c("game_id", "away_team_abbrev"))

kills_wins <- kills_wins %>%
  left_join(
    game_kills %>% select(game_id, home_team_abbrev, away_team_abbrev, home_kills, away_kills),
    by = "game_id"
  ) %>%
  mutate(
    opp_kills = case_when(
      team_abbreviation == home_team_abbrev ~ away_kills,
      team_abbreviation == away_team_abbrev ~ home_kills,
      TRUE ~ NA_real_
    ),
    kills_differential = kills - opp_kills
  )

# ============================================================
# PART 2: Logistic Regression (full league)
# ============================================================

# Simple logistic: kills ~ wins
kills_model_full <- glm(
  team_winner ~ kills,
  data = kills_wins,
  family = binomial
)
summary(kills_model_full)

# With kills differential (stronger signal)
kills_diff_model <- glm(
  team_winner ~ kills_differential,
  data = kills_wins,
  family = binomial
)
summary(kills_diff_model)

# Mixed effects model (accounts for team-level variation)
kills_mixed <- glmer(
  team_winner ~ kills + (1 | team_abbreviation),
  data = kills_wins,
  family = binomial
)
summary(kills_mixed)

# ============================================================
# PART 3: Miller's 7+ Kill Threshold Test
# ============================================================

kills_wins <- kills_wins %>%
  mutate(seven_plus_kills = kills >= 7)

threshold_summary <- kills_wins %>%
  group_by(seven_plus_kills) %>%
  summarise(
    win_rate = mean(win_num, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

print(threshold_summary)

# Win rate across kill buckets
kill_bucket_summary <- kills_wins %>%
  mutate(kill_bucket = cut(kills, breaks = c(-Inf, 2, 4, 6, 8, 10, Inf),
                            labels = c("0-2", "3-4", "5-6", "7-8", "9-10", "11+"))) %>%
  group_by(kill_bucket) %>%
  summarise(
    win_rate = mean(win_num, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# ============================================================
# PART 4: Distribution of Kills per Game
# ============================================================

p_dist <- ggplot(kills_wins, aes(x = kills)) +
  geom_histogram(binwidth = 1, fill = "#1D428A", color = "white") +
  geom_vline(xintercept = 7, color = "#C8102E", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 7.4, y = Inf, label = "Miller's 7-kill threshold",
           hjust = 0, vjust = 1.5, color = "#C8102E", size = 3.5) +
  labs(
    title = "Distribution of Kills per Game — Full NBA 2024-25",
    x = "Kills", y = "Count"
  ) +
  theme_minimal()

# ============================================================
# PART 5: Win Rate by Kills — Full League
# ============================================================

league_kill_summary <- kills_wins %>%
  group_by(kills) %>%
  summarise(
    win_rate = mean(win_num, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 10)  # filter sparse kill counts

p_winrate <- ggplot(league_kill_summary, aes(x = kills, y = win_rate, size = n)) +
  geom_point(color = "#1D428A", alpha = 0.8) +
  geom_smooth(aes(weight = n), method = "loess", se = TRUE,
              color = "#C8102E", fill = "#C8102E", alpha = 0.15) +
  geom_vline(xintercept = 7, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent_format()) +
  scale_size_continuous(name = "Games") +
  labs(
    title = "NBA Win Rate by Kills — Full League 2024-25",
    x = "Kills", y = "Win %"
  ) +
  theme_minimal()

# ============================================================
# PART 6: Kills Differential vs Win Rate
# ============================================================

diff_summary <- kills_wins %>%
  group_by(kills_differential) %>%
  summarise(
    win_rate = mean(win_num, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 10)

p_diff <- ggplot(diff_summary, aes(x = kills_differential, y = win_rate, size = n)) +
  geom_point(color = "#1D428A", alpha = 0.8) +
  geom_smooth(aes(weight = n), method = "loess", se = TRUE,
              color = "#C8102E", fill = "#C8102E", alpha = 0.15) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Win Rate by Kills Differential (Your Kills − Opponent Kills)",
    x = "Kills Differential", y = "Win %"
  ) +
  theme_minimal()

# ============================================================
# PART 7: Team-Level Kills Leaderboard (season averages)
# ============================================================

team_kill_avgs <- kills_wins %>%
  group_by(team_abbreviation) %>%
  summarise(
    avg_kills = mean(kills, na.rm = TRUE),
    avg_opp_kills = mean(opp_kills, na.rm = TRUE),
    avg_diff = mean(kills_differential, na.rm = TRUE),
    win_pct = mean(win_num, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_kills))

print(team_kill_avgs)

# Scatter: avg kills vs win pct by team
p_team <- ggplot(team_kill_avgs, aes(x = avg_kills, y = win_pct, label = team_abbreviation)) +
  geom_point(color = "#1D428A", size = 3) +
  geom_text(nudge_y = 0.01, size = 2.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#C8102E", fill = "#C8102E", alpha = 0.15) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Team Avg Kills vs Win % — 2024-25 NBA",
    x = "Avg Kills per Game", y = "Win %"
  ) +
  theme_minimal()

# ============================================================
# PART 8: Multi-Season Analysis (2020-2025)
# ============================================================

multi_team_box <- load_nba_team_box(seasons = 2020:2025)

# Compute kills season by season, tagging each with the season year
seasons <- 2020:2025

multi_season_kills <- map_dfr(seasons, function(s) {
  message("Loading PBP for season: ", s)
  pbp <- load_nba_pbp(seasons = s)  # load one season at a time to avoid memory issues
  
  teams <- unique(c(pbp$home_team_abbrev, pbp$away_team_abbrev))
  teams <- teams[!is.na(teams)]
  
  map_dfr(teams, function(tm) {
    compute_kills_for_team(pbp, tm) %>% mutate(season = s)
  })
})

# Join on game_id only — season is just for faceting later
multi_kills_wins <- multi_team_box %>%
  left_join(
    multi_season_kills %>% select(game_id, team_abbrev, kills, season),
    by = c("game_id", "team_abbreviation" = "team_abbrev")
  ) %>%
  mutate(
    kills   = replace_na(kills, 0),
    win_num = as.integer(team_winner)
  )

# Multi-season logistic regression
multi_model <- glm(team_winner ~ kills, data = multi_kills_wins, family = binomial)
summary(multi_model)

# Win rate by kills, faceted by season
multi_season_summary <- multi_kills_wins %>%
  group_by(season.x, kills) %>%
  summarise(win_rate = mean(win_num, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 5)

p_multi <- ggplot(multi_season_summary, aes(x = kills, y = win_rate)) +
  geom_smooth(aes(weight = n), method = "loess", se = FALSE, color = "#1D428A") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "#C8102E") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "gray60") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~season.x) +
  labs(
    title = "NBA Win Rate by Kills — 2020 to 2025",
    subtitle = "Red dashed line = Miller's 7-kill threshold",
    x = "Kills", y = "Win %"
  ) +
  theme_minimal(base_size = 12)

print(p_multi)

# ============================================================
# PART 9: Kills vs DRTG (Defensive Rating) Comparison
# ============================================================
# hoopR's team_box includes DRTG for some datasets, or you can pull from nbastatR/basketball-reference
# If available in team_box:

if ("defensive_rating" %in% colnames(kills_wins)) {
  drtg_model <- glm(
    team_winner ~ defensive_rating,
    data = kills_wins,
    family = binomial
  )
  
  kills_vs_drtg <- kills_wins %>%
    filter(!is.na(kills), !is.na(defensive_rating)) %>%
    summarise(
      kills_AIC = AIC(kills_model_full),
      drtg_AIC  = AIC(drtg_model)
    )
  
  print(kills_vs_drtg)  # Lower AIC = better fit
}

# ============================================================
# PRINT ALL PLOTS
# ============================================================

print(p_dist)
print(p_winrate)
print(p_diff)
print(p_team)
print(p_multi)
```




